R Notebook

### 2017-04-11 doc/RMD/data-comparison.Rmd
update data-comparison to filter data as well

```{r}
size_ranges <- 
"locus\trange\tn
5-2	318-324	4
6-2	483-495	3
7-2	158-174	7
8-3	244-270	7
9-2	360-382	9
12-2	214-222	5
17-3	342-363	7
20-3	280-282	2
55-4	153-216	10
110-4	370-386	5
114-4	339-416	10
"
sr <- read_tsv(size_ranges) %>%
  separate(range, c("lower", "upper"), sep = "-") # split range into lower and upper bounds
sr
# Making the assertation string.
# To do this, we are using `sprintf()` to print the assert statement:
#   assert(within_bounds(lower, upper), `locus`)
# which means: "assert that the alleles (data) within this locus (variable) are
# within the defined range". 
# 
# The `apply()` function takes a matrix or data frame and applies a given function over each
# row (MARGIN = 1) or column (MARGIN = 2). In this case, we are specifying rows.
# 
# We then collapse all these statements with the pipe operator: %>%
# Then we add the name of the operation to the beginning
assrt_string <- apply(sr, MARGIN = 1, function(x){
    sprintf("  assert(within_bounds(%s, %s), `%s`)", x[2], x[3], x[1]) 
  }) %>% 
  paste(collapse = " %>%\n") %>%
  paste("assert_alleles_within_bounds <- .", ., sep = " %>%\n")
#
# now we can see what this statement looks like
cat(assrt_string)
# here, we evaluate it, making it available for us later on.
eval(parse(text = assrt_string))

@@

```{r sydney_data}
syd <- read_excel("../Analysis4 ForManu/A1_Copy of binned-genotypes_SE.xlsx", 
                sheet = "GenAlexBinned", skip = 1) %>%
  select(-1) %>%                # removing first column, which is empty
  gather(locus, allele, -1) %>% # gather all loci into tidy columns
  mutate(locus = trimws(locus) %>% substr(1, nchar(.) - 3)) %>% # remove (F) designator
  mutate(allele = as.integer(allele)) %>% # force alleles to integers
  spread(locus, allele) %>%     # spread data out with individual loci in columns
  separate(iso_st_mcg_org_loc_yr_hst_cult_rep, 
           c("Isolate", "Severity", "MCG", "State", "Source", "Year", "Host"), 
           sep = "_") %>%
  mutate_if(is.character, trimws) %>% # Trim whitespace forom all character columns
  mutate(Severity = as.numeric(Severity)) %>%
  mutate(Source = ifelse(Source == "", "unk", Source)) %>%
  slice(-n()) %>% # remove last row
  arrange(Isolate) %>%
  check_data_cromulence
syd
```

@@

```{r sajeewa_data}
column_specification <- cols(
  Individual = col_character(),
  iso_st_mcg_org_loc_yr_hst = col_character(),
  `5-2(F)` = col_integer(),
  `6-2(F)` = col_integer(),
  `7-2(F)` = col_integer(),
  `8-3(H)` = col_integer(),
  `9-2(F)` = col_integer(),
  `12-2(H)` = col_integer(),
  `17-3(H)` = col_integer(),
  `20-3(F)` = col_integer(),
  `55-4(F)` = col_integer(),
  `110-4(H)` = col_integer(),
  `114-4(H)` = col_integer()
)
saj <- read_csv("../Analysis4 ForManu/A2_Copy4 EUR_AUS_forManu.csv", skip = 2, col_types = column_specification) %>%
  select(-1) %>%
  gather(locus, allele, -1) %>% # gather all loci into tidy columns
  mutate(locus = trimws(locus) %>% substr(1, nchar(.) - 3)) %>% # remove (F) designator
  spread(locus, allele) %>%     # spread data out with individual loci in columns
  separate(iso_st_mcg_org_loc_yr_hst, # Note: this corresponds closer to the initial data
           c("Isolate", "Severity", "MCG", "State", "Source", "Year", "Host"), 
           sep = "_") %>%
  mutate_if(is.character, trimws) %>% # Trim whitespace forom all character columns
  mutate(Severity = as.numeric(Severity)) %>%
  arrange(Isolate) %>%
  check_data_cromulence
saj
```

@@

>>>>

```{r}
size_ranges <- 
"locus\trange\tn
5-2(F)	318-324	4
6-2(F)	483-495	3
7-2(F)	158-174	7
8-3(H)	244-270	7
9-2(F)	360-382	9
12-2(H)	214-222	5
17-3(H)	342-363	7
20-3(F)	280-282	2
55-4(F)	153-216	10
110-4(H)	370-386	5
114-4(H)	339-416	10
"
sr <- read_tsv(size_ranges) %>%
  separate(range, c("lower", "upper"), sep = "-") # split range into lower and upper bounds
sr
# Making the assertation string.
# To do this, we are using `sprintf()` to print the assert statement:
#   assert(within_bounds(lower, upper), `locus`)
# which means: "assert that the alleles (data) within this locus (variable) are
# within the defined range". 
# 
# The `apply()` function takes a matrix or data frame and applies a given function over each
# row (MARGIN = 1) or column (MARGIN = 2). In this case, we are specifying rows.
# 
# We then collapse all these statements with the pipe operator: %>%
# Then we add the name of the operation to the beginning
assrt_string <- apply(sr, MARGIN = 1, function(x){
    sprintf("  assert(within_bounds(%s, %s), `%s`)", x[2], x[3], x[1]) 
  }) %>% 
  paste(collapse = " %>%\n") %>%
  paste("assert_alleles_within_bounds <- .", ., sep = " %>%\n")
#
# now we can see what this statement looks like
cat(assrt_string)
# here, we evaluate it, making it available for us later on.
eval(parse(text = assrt_string))

@@

```{r sydney_data}
syd <- read_excel("../Analysis4 ForManu/A1_Copy of binned-genotypes_SE.xlsx", 
                sheet = "GenAlexBinned", skip = 1) %>%
  select(-1) %>%                # removing first column, which is empty
  gather(locus, allele, -1) %>% # gather all loci into tidy columns
  mutate(locus = trimws(locus)) %>% # remove whitespace in locus names
  mutate(allele = as.integer(allele)) %>% # force alleles to integers
  spread(locus, allele) %>%     # spread data out with individual loci in columns
  separate(iso_st_mcg_org_loc_yr_hst_cult_rep, 
           c("Isolate", "Severity", "MCG", "State", "Source", "Year", "Host"), 
           sep = "_") %>%
  mutate_if(is.character, trimws) %>% # Trim whitespace forom all character columns
  mutate(Severity = as.numeric(Severity)) %>%
  mutate(Source = ifelse(Source == "", "unk", Source)) %>%
  slice(-n()) %>% # remove last row
  arrange(Isolate) %>%
  check_data_cromulence
syd
```

@@

```{r sajeewa_data}
column_specification <- cols(
  Individual = col_character(),
  iso_st_mcg_org_loc_yr_hst = col_character(),
  `5-2(F)` = col_integer(),
  `6-2(F)` = col_integer(),
  `7-2(F)` = col_integer(),
  `8-3(H)` = col_integer(),
  `9-2(F)` = col_integer(),
  `12-2(H)` = col_integer(),
  `17-3(H)` = col_integer(),
  `20-3(F)` = col_integer(),
  `55-4(F)` = col_integer(),
  `110-4(H)` = col_integer(),
  `114-4(H)` = col_integer()
)
saj <- read_csv("../Analysis4 ForManu/A2_Copy4 EUR_AUS_forManu.csv", skip = 2, col_types = column_specification) %>%
  select(-1) %>%
  gather(locus, allele, -1) %>% # gather all loci into tidy columns
  mutate(locus = trimws(locus)) %>% # remove any whitespace in the locus names
  spread(locus, allele) %>%     # spread data out with individual loci in columns
  separate(iso_st_mcg_org_loc_yr_hst, # Note: this corresponds closer to the initial data
           c("Isolate", "Severity", "MCG", "State", "Source", "Year", "Host"), 
           sep = "_") %>%
  mutate_if(is.character, trimws) %>% # Trim whitespace forom all character columns
  mutate(Severity = as.numeric(Severity)) %>%
  arrange(Isolate) %>%
  check_data_cromulence
saj
```

@@

## Saving data as genclone object

First we need to transform the data to something that

```{r}
dat <- saj %>% 
  select(-(Isolate:Host)) %>% 
  df2genind(ind.names = saj$Isolate, 
            strata = select(saj, MCG:Host), # Filtering out severity and Isolate
            ploidy = 1) %>%
  as.genclone()
nameStrata(dat)[2] <- "Region"
pops <- levels(strata(dat)$Region)
levels(strata(dat)$Region) <- case_when(pops == "FR" ~ "France", 
                                        pops == "MX" ~ "Mexico", 
                                        pops == "AU" ~ "Australia", 
                                        TRUE ~ pops)
dat
locNames(dat)

repeat_lengths <-
  c(
  `5-2(F)` = 2.000000,
  `5-3(F)` = 0.500000,
  `6-2(F)` = 6.000000,
  `7-2(F)` = 2.000000,
  `8-3(H)` = 2.000000,
  `9-2(F)` = 2.000000,
  `12-2(H)` = 2.000000,
  `17-3(H)` = 3.000000,
  `20-3(F)` = 2.000000,
  `36-4(F)` = 0.500000,
  `50-4(F)` = 0.500000,
  `55-4(F)` = 4.000000,
  `92-4(F)` = 2.000000,
  `106-4(H)` = 4.000000,
  `110-4(H)` = 4.000000,
  `114-4(H)` = 4.000000
  )
repeat_lengths <- ifelse(repeat_lengths < 1, 4, repeat_lengths)
```

### Adding inconsistent loci

Only 11 loci were manually curated by Sydney and Sajeewa. I'm taking a look at
the other 5 loci because they may be informative. In order to do that, I'm going
to have to clean them up by estimating the the true allele size.

```{r read_excel}
ex <- readxl::read_excel("../Analysis4 ForManu/A1_Copy of binned-genotypes_SE.xlsx", sheet = "GenAlex", skip = 1) %>%
  select(-1) %>%                # removing first column, which is empty
  gather(locus, allele, -1) %>% # gather all loci into tidy columns
  mutate(locus = trimws(locus)) %>% # remove (F) designator
  mutate(allele = as.integer(allele)) %>% # force alleles to integers
  spread(locus, allele) %>%
  slice(-n())
ex <- ex[!names(ex) %in% locNames(dat)]

# Function to select an adjacent allele. It will select the
# next allele if the next allele is not missing and it's distance
# is one away and the previous allele for the same conditions.
# If none of the conditions are met, it will retain the allele.
cromulent_allele <- Vectorize(function(lower, allele, higher){
  if (!is.na(higher) && abs(allele - higher) == 1){
    out <- higher
  } else if (!is.na(lower) && abs(allele - lower) == 1){
    out <- lower
  } else {
    out <- allele
  }
  out
})
ex
exsummary <- ex %>% 
  gather(locus, allele, -1) %>% # tidy the data
  group_by(locus, allele) %>%   
  summarize(n = n()) %>%        # summarize by count 
  ungroup() %>%
  group_by(locus) %>%           # group the loci, add the lower and upper alleles,
  mutate(lower = lag(allele), higher = lead(allele)) %>% # and then create new_alleles
  mutate(new_allele = ifelse(n < 3, cromulent_allele(lower, allele, higher), allele)) %>%
  select(locus, new_allele, allele)
exsummary
```

Now that we have our data set filtered (to a degree), we can merge the data with
`dat` that we defined above.

```{r merging_data}
corrected_loci <- ex %>% gather(locus, allele, -1) %>%
  left_join(exsummary, by = c("locus", "allele")) %>%
  mutate(allele = new_allele) %>%
  select(-new_allele) %>%
  spread(locus, allele) %>%
  separate(iso_st_mcg_org_loc_yr_hst_cult_rep, # Note: this corresponds closer to the initial data
           c("Isolate", "Severity", "MCG", "State", "Source", "Year", "Host"), 
           sep = "_") %>%
  arrange(Isolate) %>%
  select(-(Severity:Host))
datdf <- genind2df(dat, usepop = FALSE) %>% 
  rownames_to_column(var = "Isolate") %>% 
  left_join(corrected_loci, by = "Isolate")
stopifnot(identical(datdf$Isolate, indNames(dat)))
dat   <- datdf %>% 
  select(-Isolate) %>% 
  df2genind(, ind.names = indNames(dat), strata = strata(dat), ploidy = 1) %>% 
  as.genclone()
```

The original data includes both Severity and Isolate. Since these are not
necessary for delimiting the strata, we will place them in the "other" slot
after converting Severity to numeric. Placing this information in the "other"
slot ensures that these data will travel with the object.

```{r fix_strata}
stopifnot(identical(indNames(dat), saj$Isolate))
other(dat)$meta <- select(saj, Severity, Isolate)
other(dat)$REPLEN <- fix_replen(dat, repeat_lengths)
```

```{r, save_data}
save(dat, datdf, corrected_loci, file = "data/sclerotinia_16_loci.dat")
```
